Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support)

### 2017-02-14 csf-2017/paper.tex
added sentence about stack being abstraction on memory
Some notes and small changes to the LR section.

\section{Logical relation}
\label{sec:logical-relation}
In order to be able to reason about programs running on the capability
machine and prove properties such as well-bracketedness and
encapsulation about them, we define a unary logical relation.
% Intuition
The logical relation is a semantic model of all the words on the
capability machine that are definitely safe to give to an
adversary. Whether a capability is safe to give to an adversary
depends heavily on the contents of the memory since instructions as well
as capabilities may reside in memory. Hence we index the logical
relation over worlds, which are used to model the memory, and our logical
relation is thus an example of what is often referred to as a Kripke
logical relation. A world consists of a finite set of regions, and each region
can be understood as a protocol describing the possible evolution of
part of the memory. Memory changes during an
execution since new memory can be allocated and part of the existing
memory can change. We use two future world relations to capture
such changes to the memory in our model.
%

\subsection{Worlds}
%\todo[inline]{Find a good place to describe monotonicity of
%interpretation function.}
% \lau{this may not be an important enough point to include it.}
\begin{figure}
  \centering
  \includegraphics{w11}
  \caption{Illustration of the relation between local and global capabilities and temporary and permanent regions. The table represents a memory and the colored fields are regions governing parts of the memory. Roughly speaking, global capabilities should never depend on memory governed by temporary regions.}
  \label{fig:cap-world}
\end{figure}
A world is a finite map from region names, modeled as natural numbers, to regions.
We have three types of regions: \emph{permanent}, \emph{temporary}, and \emph{revoked}.
%$\perma$, $\temp$, and $\revoked$. The
%$\perma$ and $\temp$ 
Each permanent and temporary region contains a state transition system, with
public and private transitions, which describes the protocol governing
the evolution of the memory described by the region, similarly to what
has been used in logical relations for high-level
languages~\cite{Ahmed:popl09,Dreyer:jfp12,Devriese:2016ObjCap}. 
The protocols imposed on memory by permanent regions stay in place
indefinitely. This means that any capability, local or global, can
depend on these protocols. The protocols imposed on memory by temporary
regions can be revoked at certain points during execution,
corresponding to the points when local capabilities can be
revoked. This means that local capabilities safely can depend on the
protocols imposed on memory by temporary regions, whereas global
capabilities cannot, since a global capability may stay around
throughout the execution of a program, but a temporary region may be
revoked. This is illustrated in \figurename~\ref{fig:cap-world}.

For technical reasons (monotonicity of the logical relation), a
temporary region is not removed in the model when it is
revoked. Instead it is turned into a new kind of region, namely a
revoked region. A revoked region contains no state transition
system and puts no requirements on the memory it models. It
simply serves as a mask for a temporary region that have been revoked.
This way of masking a region goes back to earlier work of
Ahmed~\cite{Ahmed2004semantics} and was also used in~\cite{Thamsborg:2011:KLR:2034773.2034831}.

@@

The worlds end up being defined recursively: Worlds describe memory,
which can contain capabilities, whose safety depends on worlds. 
We use the method of Birkedal et
al.~\cite{Birkedal:2011:SKM:1926385.1926401} 
for constructing the worlds.

% Short technical explanation
% The recursiveness of the worlds gives rise to a recurisve domain
% equation. Theorem~\ref{thm:world-existence} gives the solution to this
% recursive domain equation.
\begin{theorem}\label{thm:world-existence}
  There exists a \cofe{} $\Wor$ and preorders $\futurestr$ and
  $\futurewk$ such that $(\Wor,\futurestr)$ and $(\Wor,\futurewk)$ are
  preordered \cofes{}, and there exists an isomorphism $\xi$ such that
  {\small
    \begin{align*}              
      \xi : \Wor & \cong \blater (\nats \finparfun \Regions)\\
      \Regions & = \{\revoked\} \uplus \\
                 &\;\quad \begin{multlined}
                     \{\temp\} \times \States \times \Rels \times \\
                     (\States \fun (\Wor \monwknefun \UPred{\HeapSegments}))
                   \end{multlined} \uplus \\
                 &\;\quad \begin{multlined}
                     \{\perma\} \times \States \times \Rels \times \\
                     (\States \fun (\Wor \monstrnefun \UPred{\HeapSegments}))
                   \end{multlined}
    \end{align*}
  } and for $W, W' \in \Wor$
  \begin{align*}
    W' \futurestr W & \Leftrightarrow \xi(W') \futurestr \xi(W)   \\
    W' \futurewk W & \Leftrightarrow \xi(W') \futurewk \xi(W)
  \end{align*}
\end{theorem}
We use the $\Regions$ from Theorem~\ref{thm:world-existence} to define
the worlds:
\begin{align*}
  \Worlds & = \RegionNames \finparfun \Regions\\
 \RegionNames & = \nats
\end{align*}

@@

The properties of the \emph{private future region} relation are:
\begin{mathpar}
  \inferrule{ (v,\phi_\pub,\phi,H) = (v',\phi_\pub',\phi',H') \\
              (s,s') \in \phi} 
  {  (v',s',\phi_\pub',\phi',H') \futurestr (v,s,\phi_\pub,\phi,H) }
  \and
  \inferrule{ r \in \Regions }
  { r \futurestr (\temp,s,\phi_\pub,\phi,H) }
  \and
  \inferrule{ r \in \Regions }
  { r \futurestr \revoked }
\end{mathpar}
These rules express that, in the private future region relation, 
we allow revocation of temporary regions. 
As explained above, this can be done using a special revoked region,
which does not describe any memory, but we also allow  
any other region to serve as a mask, and thus any region $r$ can be a
future region of a temporary region and of a revoked region.
If a temporary region is not masked by a region, then it behaves like a
permanent region. The permanent regions stay permanent and they are
allowed to take a transition in the private part of the transition
system.

The \emph{public future} region relations satisfies the following properties:
\begin{mathpar}
  \inferrule{ (v,\phi_\pub,\phi,H) = (v',\phi_\pub',\phi',H') \\
    (s,s') \in \phi_\pub }
  {  (v',s',\phi_\pub',\phi',H') \futurewk (v,s,\phi_\pub,\phi,H) }
  \and
  \inferrule{ (\temp,s,\phi_\pub,\phi,H) \in \Regions }
  { (\temp,s,\phi_\pub,\phi,H) \futurewk \revoked }
  \and
  \inferrule{ }
  { \revoked \futurewk \revoked }
\end{mathpar}
In the public future regions we do not permit revocation of temporary
regions. Instead, temporary and permanent regions are only 
allowed to take a transition in the public part of the transition system.
In the public future region relation, we only allow revoked and
temporary regions to mask a region. This means that the private future
world relations allows us to reuse a region that has been revoked
earlier.

@@

\caption{Permission-based conditions}
\label{fig:perm-conds}
\end{figure}
}
%subsection introduction
The logical relation defines semantically what it means to be
capability safe. In large parts, capability safety is defined by
when a capability safely can be given to an untrusted piece of
code. In the following paragraphs, we will go through the definitions
of the logical relation and provide an intuition why they define
capability safety.
The definition of the logical relation is found in
\figurename~\ref{fig:logrel}, \figurename~\ref{fig:standard-regions},
and \figurename~\ref{fig:perm-conds}. Some definitions are omitted and
explained when needed. The precise definitions can be found in the
technical appendix~\cite{XXX}.

The \emph{observation relation}, $\observations$, defines which
configurations are capability safe. Roughly speaking,
a configuration is capability safe if, whenever it halts, then
the resulting memory satisfies a private future world. 
Notice that if a configuration fails, e.g., if an unauthorized access 
(e.g., a load of a memory location without the requisite read capability)
is attempted, then the configuration is considered capability safe, which it should
because then the capability machine has caught the attempted unauthorized access
This is similar to the logical relation given for an untyped language
in~\cite{Devriese:2016ObjCap}, but unlike logical relations for typed languages, which
typically require that computations do not fail.

The \emph{register-file relation}, $\stdrr$, specifies all the
register-files that are capability safe. Capability safe
register-files contain capability safe words (e.g.n, words in
$\stdvr$) in all registers but $\pcreg$.

The \emph{expression relation}, $\stder$, defines when a word is
capability safe as a program counter. That is the case when the word can be plugged into a
capability safe register file (i.e., a register file in $\stdrr$) and
paired with a memory satisfying the world to become a capability safe
configuration.

The \emph{value relation}, $\stdvr$, defines which words are capability
safe. A word is capability safe if it cannot be used to break the
protocols defined by the world. Data is always capability safe because
it provides no authority that can be used to break the
invariants. Capabilities gives the authority to manipulate memory and
potentially break memory protocols, so they need to satisfy certain
conditions to be capability safe. We define a condition for each kind
of permission a capability can have. These permission-based conditions
can be found in \figurename~\ref{fig:perm-conds}.

The condition for capabilities with read permission is
$\readCond{}$. It ensures that a read capability can only be used to
read capability safe words, by requiring that the region that governs
the part of memory that the capability grants authority over enforces
that all the words are in the value relation. Generally speaking,
global capabilities should only depend on permanent regions. Therefore
we use the function $\var{localityReg}(g,W)$, which projects out all
active regions when $g$ is local, but only the permanent regions when
$g$ is global.
The read condition uses one of two standard regions found
in~\ref{fig:standard-regions}. $\iota^\pwl_{\start,\addrend}$ requires
all the words in the range $[\start,\addrend]$ to be capability
safe. $\iota^\nwl_{\start,\addrend}$ requires the words to be
non-local and capability safe.

For a capability with a write permission, $\writeCond{}$ must be
satisfied for the capability's range of authority. Generally speaking,
write capabilities may be used to write any word, so the region that
governs the part of memory the write capability has authority over
must at least allow any capability safe word to be stored.
This is of course not entirely true because of the distinction between
write and permit-write-local capabilities. For write capabilities, the
region must require the contents of memory to be non-local (and
capability safe). For permit-write-local capabilities, the
non-local requirement is omitted.
The write condition uses $\var{localityReg}$ to pick an appropriate
region in the same way the read condition does.
The condition for the region to be address-stratified is there for
technical reasons.

The conditions $\entryCond{}$ and $\execCond{}$ are very similar and
differ only slightly. Executable capabilities can be updated to point
anywhere in memory, so they must be capability safe as a program
counter (in the $\stder$-relation) no matter where they point in memory. In contrast, enter
capabilities are opaque and must be used from the address they point
to. They change permission when jumped to, so we require them to be
capability safe as a program counter when the permission is changed to
read/execute.
In both cases, it must be safe to jump to the capability. This
means that the capability must be capability safe as a
program counter in a future world. If the capability is global,
then it should only depend on permanent regions and therefore
we require that it should be capability safe as a program counter in any \emph{private}
future world (recall that temporary regions may be revoked in a private future
world). On the other hand, a local capability should be capability
safe in any \emph{public} future world because it may depend on temporary
regions, so we cannot expect them to function correctly as program
counters if temporary regions are revoked.

The definition of the value relation in \figurename~\ref{fig:logrel}
is defined in terms of the permission-based conditions.

>>>>

\section{Logical relation}
\label{sec:logical-relation}
In order to be able to reason about programs running on the capability
machine and prove properties such as well-bracketedness and
encapsulation, we define a unary logical relation.
% Intuition
It is a semantic model of all the words on the capability machine that are
definitely safe to give to an adversary. Whether a capability is safe to give to
an adversary depends heavily on the contents of the memory since instructions as
well as capabilities may reside in memory. Hence we index the logical relation
over worlds, which are used to model the memory (making our logical relation a
Kripke logical relation). A world consists of a finite set of regions, and each
region can be understood as a protocol describing the possible evolution of part
of the memory. Future evolutions of worlds (e.g. new protocols established over
memory received from malloc, or steps in a protocol already established for a
part of memory) are modeled using a future world relation. In fact, as we will
explain below, there are two such relations to accomodate reasoning about local
capabilities.
%

\subsection{Worlds}
\dominique{Perhaps we could first try to convey the intuition of a permanent
  world with a temporary overlay, with global capabilities ignoring the overlay.
  Perhaps we could add a depiction of the world to Figure 5?}

%\todo[inline]{Find a good place to describe monotonicity of
%interpretation function.}
% \lau{this may not be an important enough point to include it.}
\begin{figure}
  \centering
  \includegraphics{w11}
  \caption{Illustration of the relation between local and global capabilities and temporary and permanent regions. The table represents a memory and the colored fields are regions governing parts of the memory. Roughly speaking, global capabilities should never depend on memory governed by temporary regions.}
  \label{fig:cap-world}
\end{figure}
A world is a finite map from region names, modeled as natural numbers, to regions.
We have three types of regions: \emph{permanent}, \emph{temporary}, and \emph{revoked}.
%$\perma$, $\temp$, and $\revoked$. The
%$\perma$ and $\temp$ 
Each permanent and temporary region contains a state transition system, with
public and private transitions, which describes the protocol governing
the memory described by the region, similarly to what
has been used in logical relations for high-level
languages~\cite{Ahmed:popl09,Dreyer:jfp12,Devriese:2016ObjCap}. 
The protocols imposed on memory by permanent regions stay in place
indefinitely. This means that any capability, local or global, can
depend on these protocols. The protocols imposed on memory by temporary
regions can be revoked at certain points during execution,
corresponding to the points, on the condition that we clear local capabilities can be
revoked. This means that local capabilities safely can depend on the
protocols imposed on memory by temporary regions, whereas global
capabilities cannot, since a global capability may stay around
throughout the execution of a program, but a temporary region may be
revoked. This is illustrated in \figurename~\ref{fig:cap-world}.

For technical reasons, a temporary region is not removed in the model when it is
revoked. Instead it is turned into a new kind of region, namely a revoked
region. A revoked region contains no state transition system and puts no
requirements on the memory it models. It simply serves as a mask for a
temporary region that has been revoked. This way of masking a region goes back
to earlier work of \citet{Ahmed2004semantics} and was also used
by \citet{Thamsborg:2011:KLR:2034773.2034831}.

@@

\dominique{``end up being defined recursively'' does not sound very insightful}
The worlds end up being defined recursively: Worlds describe memory, which can
contain capabilities, whose safety depends on worlds. We use the method of
\citet{Birkedal:2011:SKM:1926385.1926401} for constructing the worlds.

% Short technical explanation
% The recursiveness of the worlds gives rise to a recurisve domain
% equation. Theorem~\ref{thm:world-existence} gives the solution to this
% recursive domain equation.
\dominique{it is not explained what cofes are?}
\begin{theorem}\label{thm:world-existence}
  There exists a \cofe{} $\Wor$ and preorders $\futurestr$ and
  $\futurewk$ such that $(\Wor,\futurestr)$ and $(\Wor,\futurewk)$ are
  preordered \cofes{}, and there exists an isomorphism $\xi$ such that
  {\small
    \begin{align*}              
      \xi : \Wor & \cong \blater (\nats \finparfun \Regions)\\
      \Regions & = \{\revoked\} \uplus \\
                 &\;\quad \begin{multlined}
                     \{\temp\} \times \States \times \Rels \times \\
                     (\States \fun (\Wor \monwknefun \UPred{\HeapSegments}))
                   \end{multlined} \uplus \\
                 &\;\quad \begin{multlined}
                     \{\perma\} \times \States \times \Rels \times \\
                     (\States \fun (\Wor \monstrnefun \UPred{\HeapSegments}))
                   \end{multlined}
    \end{align*}
  } and for $W, W' \in \Wor$
  \begin{align*}
    W' \futurestr W & \Leftrightarrow \xi(W') \futurestr \xi(W)   \\
    W' \futurewk W & \Leftrightarrow \xi(W') \futurewk \xi(W)
  \end{align*}
\end{theorem}
We use the $\Regions$ from Theorem~\ref{thm:world-existence} to define
the worlds:
\begin{align*}
  \Worlds & = \RegionNames \finparfun \Regions\\
 \RegionNames & = \nats
\end{align*}

@@

The properties of the \emph{private future region} relation are:
\begin{mathpar}
  \inferrule{ (v,\phi_\pub,\phi,H) = (v',\phi_\pub',\phi',H') \\
              (s,s') \in \phi} 
  {  (v',s',\phi_\pub',\phi',H') \futurestr (v,s,\phi_\pub,\phi,H) }
  \and
  \inferrule{ r \in \Regions }
  { r \futurestr (\temp,s,\phi_\pub,\phi,H) }
  \and
  \inferrule{ r \in \Regions }
  { r \futurestr \revoked }
\end{mathpar}
These rules express that, in the private future region relation, we allow
revocation of temporary regions. As explained above, this can be done using a
special revoked region, which does not describe any memory, but we also allow
any other region to serve as a mask, and thus any region $r$ can be a future
region of a temporary region and of a revoked region. On the other hand,
permanent regions cannot be masked away. They are only allowed to take a
transition in the private part of the transition system.

The \emph{public future} region relation satisfies \dominique{is defined by?} the following properties:
\begin{mathpar}
  \inferrule{ (v,\phi_\pub,\phi,H) = (v',\phi_\pub',\phi',H') \\
    (s,s') \in \phi_\pub }
  {  (v',s',\phi_\pub',\phi',H') \futurewk (v,s,\phi_\pub,\phi,H) }
  \and
  \inferrule{ (\temp,s,\phi_\pub,\phi,H) \in \Regions }
  { (\temp,s,\phi_\pub,\phi,H) \futurewk \revoked }
  \and
  \inferrule{ }
  { \revoked \futurewk \revoked }
\end{mathpar}
In the public future regions we do not permit revocation of temporary regions.
Instead, temporary and permanent regions are only allowed to take a transition
in the public part of the transition system. Revoked regions can either remain
revoked or be masked away by a temporary region. This means that the public
future world relations allows us to reuse a region that has been revoked
earlier.

@@

\caption{Permission-based conditions}
\label{fig:perm-conds}
\end{figure}
}
%subsection introduction
\dominique{I would try to motivate the LR defs by saying that the LR is
  constructed to express the FTLR and the FTLR says that if a piece of code only
  has access to values respecting the invariants in a world, then it will
  respect them itself too.} The logical relation defines semantically what it
means to be capability safe.\dominique{We have not introduced capability
  safety..} In large parts, capability safety is defined by when a capability
can be safely given to an untrusted piece of code. In the following paragraphs,
we will go through the definitions of the logical relation and provide an
intuition why they define capability safety. The definition of the logical
relation is found in \figurename~\ref{fig:logrel},
\figurename~\ref{fig:standard-regions}, and \figurename~\ref{fig:perm-conds}.
Some definitions are omitted and explained when needed. The precise definitions
can be found in the technical appendix~\cite{XXX}.

The \emph{observation relation} $\observations$ defines which configurations
are capability safe. Roughly speaking, a configuration is capability safe if,
whenever it halts, then the resulting memory satisfies a private future world.
Notice that failing is considered safe behavior and in fact, the machine will
often resort to failing when, for example, an unauthorized
access is attempted, like a load of a memory location through a
capability without read permission. This is similar to the logical relation
given for an untyped language in~\cite{Devriese:2016ObjCap}, but unlike logical
relations for typed languages, which typically require that computations do not
fail.

The \emph{register-file relation} $\stdrr$ specifies that register-files are
safe when they contain safe words (i.e., words in $\stdvr$) in all registers but
$\pcreg$. The \emph{expression relation} $\stder$ defines when a word is safe
to use as a program counter. That is the case when the word can be plugged into
a capability safe register file (i.e., a register file in $\stdrr$) and paired
with a memory satisfying the world to become a capability safe configuration.

The \emph{value relation} $\stdvr$ defines when words are safe, i.e. when they
cannot be used to break the protocols defined by the world. Non-capability data
is always safe because it provides no authority. Capabilities give the authority
to manipulate memory and potentially break memory protocols, so they need to
satisfy certain conditions to be capability safe. In
\figurename~\ref{fig:perm-conds}, we define such a condition for each kind of
permission a capability can have (we will call these permission-based
conditions).

For capabilities with read permission, the $\readCond{}$ ensures that it can
only be used to read capability safe words, by requiring that the region
governing the relevant part of memory, guarantees that the memory contains only
words in the value relation. Generally speaking, global capabilities should only
depend on permanent regions. Therefore we use the function
$\var{localityReg}(g,W)$, which projects out all active regions when $g$ is
local, but only the permanent regions when $g$ is global. The read condition
uses $\iota^\pwl_{\start,\addrend}$, a standard regions defined in
Figure~\ref{fig:standard-regions}, which requires all the words in the range
$[\start,\addrend]$ are safe.

For a capability with write permission, $\writeCond{}$ must be satisfied for the
capability's range of authority. Such a capability may be used to write any
word, so the region governing the relevant part of memory must (at least) allow
any capability safe word to be stored. However, this is not entirely true
because of the distinction between write and permit-write-local capabilities.
Only permit-write-local capabilities need to allow arbitrary safe values to be
stored, while for write capabilities, it is sufficient to allow storing values
that are both non-local and safe. The write condition uses $\var{localityReg}$
to pick an appropriate region for its locality in the same way the read
condition does. The condition for the region to be address-stratified is there
for technical reasons.

The conditions $\entryCond{}$ and $\execCond{}$ are very similar and differ only
slightly. In both cases, we require that the capability can be safely jumped to.
However, executable capabilities can be updated to point anywhere in memory, so
they must be capability safe as a program counter (in the $\stder$-relation) no
matter where they point to. In contrast, enter capabilities are opaque and can
only be used to jump to the address they point to. They also change permission
when jumped to, so we require them to be capability safe as a program counter
when the permission is changed to read/execute. Because the capabilities are not
necessarily invoked immediately, we need to require that this is the case in a
future world, but it depends on the locality of the capability which future
worlds we consider. If the capability is global, then it should only depend on
permanent regions, so we require capability safety as a program counter in any
\emph{private} future world (recall that temporary regions may be revoked in a
private future world). For local capabilities, it suffices to be capability safe
in \emph{public} future worlds, where temporary regions are still there.
